Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-26 Manuscript.qmd
added nacella text

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: 
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

@@

\##########

@@

@@

```{r Nacella_Correlation_Graphs}
#| fig-cap: Correlation graphs for *Nacella* sp. specimens
#| fig-height: 15
#| fig-width: 15

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
  # facet_wrap(~region,ncol=2,scales = "free")

@@

P_all/P_individual+plot_layout(heights = c(0.8,0.2))

>>>>

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: docx
#  elsevier-pdf:
#    keep-tex: true
#    journal:
#      name: Palaeo^3^
#      formatting: preprint
#      model: 3p
#      cite-style: authoryear
output:
  units: cm
---

@@

P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")

@@

@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their δ^18^O-values (similar graphs for the other *Nacella* sp. specimen can be found in the supplementaries). Compared to the *P. vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and its simple make-up.

@@

P1 <-
  ND_Line %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6) %>%
  filter(dist<10) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")

P2 <-
  D18O %>% 
  filter(sampleID=="ND-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))

P3 <-
  NM_Line %>% 
  filter(relstd < 0.1) %>%
  filter(dist<9.5) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")

P4 <-
  D18O %>% 
  filter(sampleID=="NM-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))

 ((P1 / P2) | (P3 / P4)) + plot_annotation(tag_levels = 'a')

@@

The *Nacella* sp. specimens showed similarly clustered correlations, which interestingly were not grouped by species. In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*Nacella magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr). Both also share a high coefficient of determination (R^2^) of 0.95 and 0.92 respectively, with the remaining specimens being somewhat lower (R^2^ = 0.76 for ND-1016-4 and R^2^=0.84 for NM-1016-3). The mean R^2^ value for each *Nacella* species is 0.86 for *Nacella deaureata* and 0.88 for *Nacella magellanica*. Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out. That said, since they both have a high R^2^ value, it might just be that they were both minimally affected by factors other than SST, including the sampling location of our line scans.

```{r Nacella_Correlation_Graphs}
#| fig-cap: Correlation graphs for *Nacella* sp. specimens
#| fig-height: 18
#| fig-width: 10
#| label: fig-Nac_Corr

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  theme_cowplot(20)+
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
 # ylim(0.05,NA)+
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12"))#+
  # facet_wrap(~region,ncol=2,scales = "free")

@@

P_individual <- P_all+ylim(0,0.4)+
  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")

### 2024-06-26 Manuscript.tex
added nacella text

Line scans reflect well the changes previously indicated by
δ\textsuperscript{18}O values and Mg/Ca intensity ratios

. That said, some of the variability of Mg/Ca intensity ratios within
one season such as the summer period of QG1-7188-1 between 0.5 and 2.0
mm distance to the shell edge (@fig-Pat\_LIBS d), is visible in a more
dramatic manner than it appears in the 2D-image, showing the downside of
line-scanning as opposed to the broader milling approach used for the
analysis of carbonate powder \citep{Ferguson2011-zl}.

@@

{\centering \includegraphics{Manuscript_files/figure-pdf/Nacella_Correlation_Graphs-1.pdf}

@@

\caption{Correlation graphs for \emph{Nacella} sp. specimens}

>>>>

Line scans reflect well the changes previously indicated by
δ\textsuperscript{18}O values and also indicate about two years of
recorded growth. That said, the 2D imaging suggests that in both
specimens a third summer can be added to the total record, which is not
captured by the line scans. The variability seen in the \emph{P.
vulgata} specimens above is not as strong, which might also be related
to the fast and consistent growth rate of the shells.

Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
NM-1016-3 in comparison to their δ\textsuperscript{18}O-values (similar
graphs for the other \emph{Nacella} sp. specimen can be found in the
supplementaries). Compared to the \emph{P. vulgata} records, these
mirror each other much more clearly, due to the brief period recorded in
the shells and its simple make-up.

@@

The \emph{Nacella} sp. specimens showed similarly clustered
correlations, which interestingly were not grouped by species. In fact,
specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3
(\emph{Nacella magellanica}) are almost identical in their fitted
equations (Figure~\ref{fig-Nac_Corr}). Both also share a high
coefficient of determination (R\textsuperscript{2}) of 0.95 and 0.92
respectively, with the remaining specimens being somewhat lower
(R\textsuperscript{2} = 0.76 for ND-1016-4 and R\textsuperscript{2}=0.84
for NM-1016-3). The mean R\textsuperscript{2} value for each
\emph{Nacella} species is 0.86 for \emph{Nacella deaureata} and 0.88 for
\emph{Nacella magellanica}. Why two specimens of different species are
so well aligned is not entirely certain and randomness cannot for now be
ruled out. That said, since they both have a high R\textsuperscript{2}
value, it might just be that they were both minimally affected by
factors other than SST, including the sampling location of our line
scans.

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.pdf}

@@

\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} sp.
specimens}
